---
title: "GSMgrids"
output:
  html_document:
    df_print: paged
---

# Introduction

Mud (silt and clay) content of surface sediments on the Norwegian continental margin is spatially predicted based on observations (response variable) and predictor variables using random forests. 

# Preparations

## Install packages

```{r packages, message=FALSE, warning=FALSE}
rm(list=ls())

library(rgdal)
library(raster)
library(stars)
library(Boruta)
library(caret)
library(usdm)
library(corrplot)
library(ggplot2)
library(sf)
library(CAST)
library(geosphere)
library(randomForest)
library(blockCV)
library(automap)
library(gstat)
library(foreach)
library(doParallel)
library(ModelMetrics)
library(forcats)
library(dplyr)
```


## Define projection and resolution

Projection based on https://projectionwizard.org/ using the AoI.
Is it possible to automate the selection of the CSR based on the AoI?

```{r projection}
#Norway
crs <- "+proj=laea +lat_0=90 +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs" 
res <- 4000
```


## Define Area of Interest (AoI)

Define the area of interest either by bounding coordinates or by loading a shapefile.

```{r aoi}
AoI <- readOGR(dsn = "input", layer = "AoI")
```


# Predictor variables

## Load raster stack with predictor variables

```{r load_predictors}
predictors <- stack("input/predictors_ngb.grd")
```


## Ensure uniform projection

Check if AoI and predictors have the defined projection. Re-project if this is not the case.

```{r}
if (crs(AoI, asText=T) != crs) {
  AoI <- spTransform(AoI, crs)
}

if (crs(predictors, asText=T) != crs) {
  predictors <- projectRaster(predictors, crs = crs, res = res)
}
```

## Limit to predictors that are relevant for mapping grain-size

```{r limit_predictors}
predictors <- predictors[[c(1,2,3,4,5,10,11,24,25)]]
names(predictors)
```



## Crop predictors to AoI

```{r crop_predictors}
predictors <- mask(crop(predictors, AoI), AoI)
plot(predictors)
```


## Minimum extent

Create an unprojected spatial polygon giving the minimum extent of all predictors. This will be used to limit the response data to those points for which predictor variable data can be extracted.

```{r min_extent}
min_ext <- sum(predictors)
min_ext[min_ext > 0] <- 1
min_ext <- rasterToPolygons(min_ext, dissolve = TRUE)
min_ext <- spTransform(min_ext, CRSobj = "+proj=longlat +datum=WGS84 +no_defs")
```


## Create a fishnet SpatialPolygons file based on predictor raster

```{r fishnet}
fishnet <- sf::as_Spatial(sf::st_as_sf(stars::st_as_stars(predictors[[1]]), as_points = FALSE, merge = FALSE))
fishnet$ID <- 1:nrow(fishnet)
```


# Response variable

## Load response

```{r load_response}
GSM_data <- read.csv("input/gsm_data.csv", header = TRUE, sep = ",")
GSM <- GSM_data[,-c(1,3,7)] #Reduce to required columns
summary(GSM)
```

## Calculate additive log-ratio

```{r calc_alr}
GSM$alrM <- log(GSM$Mud/(GSM$Gravel + GSM$Sand))
```


## Convert to spatial object

```{r convert_to_spatial}
GSM_ED50 <- subset(GSM, Datum == "ED50")
GSM_WGS84 <- subset(GSM, Datum == "WGS84")

coordinates(GSM_ED50) = ~ Lon + Lat
proj4string(GSM_ED50)<- CRS("+proj=longlat +ellps=intl +no_defs ")
GSM_ED50 <- spTransform(GSM_ED50, CRSobj = "+proj=longlat +datum=WGS84 +no_defs")

coordinates(GSM_WGS84) = ~ Lon + Lat
proj4string(GSM_WGS84)<- CRS("+proj=longlat +datum=WGS84 +no_defs")

GSM <- GSM_ED50 + GSM_WGS84
```


## Clip to minimum extent

```{r clip_response}
GSM <- crop(GSM, min_ext)
```


## Reproject

Reproject to the previously defined projection.

```{r reproject}
GSM <- spTransform(GSM, CRSobj = crs)
```


## Centroids

```{r centroids}
centroids <- st_join(st_as_sf(fishnet), st_as_sf(GSM))

centroids <- na.omit(centroids)

centroids <- centroids %>% group_by(ID) %>% summarize(alrM = mean(alrM, na.rm = TRUE))

centroids <- st_centroid(centroids)

GSM <- as(centroids, "Spatial")

plot(AoI)
plot(GSM, pch = 20, add = TRUE)

#writeOGR(GSM, dsn = "output", layer = "sample_points", driver = "ESRI Shapefile")
```


## Create a regression matrix

```{r regression matrix}
ov_gsm <- as.data.frame(extract(predictors, GSM))

rm_alrM <- cbind(GSM$alrM, ov_gsm)

names(rm_alrM)[1] <- "alrM"

summary(rm_alrM)
```


## Data exploration

### alr-Mud

```{r hist_alrM}
hist(rm_alrM$alrM, main = "", xlab = "alr-Mud")
```


# Predictor variable selection

## Boruta algorithm

Although two models will be built, it is sufficient to run the Boruta algorithm only once, as the predictors and the sample locations are identical.

```{r boruta}
set.seed(42)
B <- Boruta(rm_alrM[[1]] ~ .,data=rm_alrM[2:ncol(rm_alrM)], pValue = 0.05,
             maxRuns = 500)
B
par(mar=c(13,4,1,1), cex = 0.6)
plot(B, las=2, colCode = c("greenyellow", "yellow2", "red3", "cadetblue"), xlab = "")
```



## De-correlation analysis

To reduce redundancy in information, a de-correlation analysis is carried out. Of those predictor variables identified as important in the Boruta analysis, only those with a correlation coefficient below a set threshold are retained. However, a universally applicable threshold does not exist. Additionally, multicollinearity, i.e., collinearity between three or more variables, might exist in the data. Variance inflation factors (VIFs) are therefore additionally calculated to check for multicollinearity. As a rule of thumb, VIFs larger than 5 or 10 indicate a problematic amount of collinearity (James et al., 2017: pp. 101-102; doi: 10.1080/24754269.2021.1980261). According to Johnston et al. (2017; doi: 10.1007/s11135-017-0584-6) a VIF of 2.5 or greater is generally considered indicative of considerable collinearity.

```{r de-corr, message=FALSE, warning=FALSE}
th <- 1

repeat{
 cor_result<- vifcor(rm_alrM[rownames(subset(attStats(B), decision == "Confirmed"))], th = th,  maxobservations = nrow(rm_alrM))
 if (max(cor_result@results[,2]) >= 2.5){
   th <- th - 0.01
 } else {
   break
 }
}

max(cor_result@results[,2])
cor_result

sel_preds <- cor_result@results$Variables
seldata <- rm_alrM[c("alrM", sel_preds)]
```


##  Correlation plot

```{r correlation_plot}
corrplot.mixed(cor(rm_alrM[sel_preds]), lower.col =  "black", tl.pos = "lt", number.cex = 0.6)
```


## Environmental space

A visual check to what extent the samples cover the environmental space. This is useful as legacy data were used and no formal sampling design was applied in the analysis.

* Blue: Samples

* Grey: Environmental data (based on random subsample)

```{r env_space}
smp <- as.data.frame(sampleRandom(x = predictors[[sel_preds]], size = 5000))

for (i in 2:ncol(seldata)) {
    
  print(ggplot() +
          geom_density(data = seldata, aes(x=seldata[,i]),
          colour="cornflowerblue", fill="cornflowerblue", alpha=0.1,size=1) +
          geom_density(data = smp, aes(x=smp[,i-1]), colour="grey",fill="grey", alpha=0.1, size=1) +
          scale_x_continuous(name = names(seldata[i])) +
          theme_bw())
        
}
```


## 2D plots of environmental space

```{r 2d_env_plots}

for (i in sel_preds[2:length(sel_preds)]) {
  
  print(ggplot() +
    geom_point(data = smp, aes(x=smp[,i], y=BATH), colour="grey", alpha=1, size=2) +
    geom_point(data = seldata, aes(x=seldata[,i], y=BATH),colour="cornflowerblue", alpha=1, size=2) +
    scale_x_continuous(name = names(seldata[i])) +
    theme_bw())
}
```



## Distances in environmental space 

```{r env_space_dist}
dist_env <- plot_geodist(st_as_sf(GSM), predictors,
                     type = "feature",
                     variables = sel_preds,
                     showPlot = FALSE)

dist_env$plot
dist_env$plot + scale_x_log10()
```


## Distances in geographic space

```{r geogr_space_dist}
dist_geogr <- plot_geodist(st_as_sf(GSM), predictors,
                     type = "geo",
                     #unit="km", # Requires CAST version >= 0.7.0
                     showPlot = FALSE)

dist_geogr$plot
dist_geogr$plot + scale_x_log10()
```


# Random Forest model

## Spatial autocorrelation range

The spatial dependence structure in the raw data is determined. Specifically, the distance (range) up to which observations are spatially autocorrelated is estimated with a variogram.

```{r spatial_autocorrelation_range}
vf <- autofitVariogram(alrM ~ 1, GSM
                       #, model = c("Sph", "Exp", "Gau", "Mat", "Ste")
                       )
plot(vf)
sar <- vf$var_model$range[2]
```



## Creating spatial blocks

Spatial blocks and folds are created. The folds will be used in a spatial k-fold cross validation. The size of the blocks is determined by the spatial autocorrelation range.

Roberts et. al. (2017) suggest that blocks should be substantially bigger than the range of spatial autocorrelation (in model residual) to obtain realistic error estimates, while a buffer with the size of the spatial autocorrelation range would result in a good estimation of error.

*Should we increase the block size? This could be gauged by looking at the geographic distances plot below. The block size might be right, when sample-to-prediction and CV distances look similar.*

```{r spatial_blocks}
m <- 2.5 # Multiplier applied to block size

spBlocks <- spatialBlock(GSM, 
                         rasterLayer = predictors[[1]], 
                         theRange = sar * m, 
                         k = 10,
                         seed = 42,
                         progress = FALSE)

#writeOGR(spBlocks$blocks, dsn = "output", layer = "spBlocks", driver = "ESRI Shapefile")
```


## Reshaping index

The output from the blocking step needs to be reshaped.

```{r reshape_index}
index <- list()
for (n in 1:spBlocks$k) {
  f <- spBlocks[["folds"]][[n]][[-2]]
  index[[length(index)+1]] <- f
}
```


## Distances in geographic space including CV distances

```{r geogr_space_dist2}
dist_geogr2 <- plot_geodist(st_as_sf(GSM), predictors,
                     cvfolds= index,
                     type = "geo",
                     #unit="km",  # Requires CAST version >= 0.7.0
                     showPlot = FALSE)

dist_geogr2$plot
dist_geogr2$plot + scale_x_log10()
```


## Model tuning

A Random Forest model is tuned. Predictor variables are finally selected in a forward feature selection approach and various values of the mtry parameter are tested in a spatial k-fold cross validation.

This step is time-consuming and memory-heavy. Therefore, only a subset of possible mtry values is tested. These are multiples of the default mtry values or the default values. 

The maximum number of iterations can be calculated upfront, based on the number of pre-selected predictors:

```{r max_iter}
factorial(length(sel_preds))/(factorial(2)*factorial(length(sel_preds)-2)) + sum(c((length(sel_preds)-2):1))
```


### Forward feature selection

The best combination of predictor variables (features) is found in a forward feature selection process.


```{r model_tuning}
nCores <- detectCores()
cl <- makePSOCKcluster(nCores - 1)
registerDoParallel(cl)

set.seed(42)

model <- ffs(seldata[sel_preds],
               seldata$alrM,
               metric = "Rsquared",
               method="rf",
               replace = FALSE,
               importance = TRUE,
               #tuneGrid = expand.grid(mtry =  round(floor(sqrt(length(sel_preds)))*c(1,1.5,2,3))),
               trControl = trainControl(method="CV",
                                        number = 10,
                                        savePredictions = "final",
                                        index = index, 
                                        allowParallel = TRUE),
               verbose = TRUE)

stopCluster(cl)

model

sel_preds <- model$selectedvars
```


### FFS plot

Plot of R2 over the model runs.

```{r ffs_plot}
plot_ffs(model)
```


## Validation statistics

The validation results of the optimal RF model.

Note that these are the statistics based on the predicted values of the selected model. These differ slightly from the values from the tuning (above), which are the means of the k predictions based on the folds.

```{r validation_stats}
t <- data.frame(model$pred$pred, model$pred$obs)

validation <- data.frame(mse=numeric(), rmse=numeric(), r2=numeric())
validation[1,1] <- round(mse(t$model.pred.obs, t$model.pred.pred), 3)
validation[1,2] <- round(rmse(t$model.pred.obs, t$model.pred.pred), 3)
validation[1,3] <- round(cor(t$model.pred.obs, t$model.pred.pred)^2, 3)

colnames(validation) <- c("MSE", "RMSE", "r2")
rownames(validation) <- NULL
validation
```

## Validation plot

```{r validation_plot}
ggplot(t, aes(x = model.pred.pred, y = model.pred.obs)) +
  geom_hex(bins = 60) +
  geom_abline(intercept = 0, slope = 1, colour = "grey", size = 1.2) +
  scale_fill_continuous(type = "viridis") +
  theme_bw() +
  scale_x_continuous(name = "Predicted value") +
  scale_y_continuous(name = "Observed value")
       
```


## Variable importance

```{r variable_importance_plot}
imp <- varImp(model$finalModel)
imp$Predictor <- rownames(imp)
rownames(imp) <- NULL
imp <- imp[order(imp[1], decreasing = TRUE), c(2, 1)]
colnames(imp)[2] <- "IncMSE"
imp

impfig <- imp %>%
  mutate(Predictor = fct_reorder(Predictor, IncMSE)) %>%
  ggplot( aes(x=Predictor, y=IncMSE)) +
    geom_bar(stat="identity", fill="#f68060", alpha=.6, width=.4) +
    coord_flip() +
    xlab("") +
    ylab("% increase in MSE") +
    theme_bw()
    
impfig
```


## Distances in environmental space including CV distances

```{r env_space_dist2}
dist_env2 <- plot_geodist(st_as_sf(GSM), predictors,
                     type = "feature",
                     cvfolds= index,
                     variables = sel_preds,
                     showPlot = FALSE)

dist_env2$plot
dist_env2$plot + scale_x_log10()
```



## Partial dependence

Partial dependence plots give a graphical depiction of the marginal effect of a variable on the response.

```{r partial_plots}
for (i in 1:length(sel_preds)) {
  partialPlot(x = model$finalModel, pred.data = seldata, x.var = sel_preds[i], main = "", xlab = sel_preds[i], ylab = "alr-Mud (-)")
}
```


# Predict RF model

## Predict mud content

```{r predict_mud}
alrM <- predict(predictors, model$finalModel)

mud <- exp(alrM)/(exp(alrM) + 1)
hist(mud, main = "", xlab = "Mud content (-)")
```


## Area of Applicability

```{r aoa}
alrM_trainDI <- trainDI(model = model,
                        variables = sel_preds)
print(alrM_trainDI)

cl <- makeCluster(nCores - 1)
registerDoParallel(cl)

alrM_aoa <- aoa(newdata = predictors, 
                model = model,
                trainDI = alrM_trainDI,
                variables = sel_preds,
                cl = cl
                )

stopCluster(cl)

plot(alrM_aoa)
```


## Plot results

```{r plot_results}
plot(mud)
plot(alrM_aoa$DI)
plot(alrM_aoa$AOA)
paste0("AOA = ", round(cellStats(alrM_aoa$AOA, stat = "mean"), 2))
```


## Convert AOA from raster to polygon

```{r aoa_poly}
aoa_poly <- rasterToPolygons(alrM_aoa$AOA, dissolve = TRUE)
plot(aoa_poly)

writeOGR(aoa_poly, dsn = "output", layer = paste0("mud_aoa_", Sys.Date()), driver = "ESRI Shapefile")
```




## Export geoTifs

```{r geotifs}
writeRaster(mud, paste0("output/mud_", Sys.Date(), ".tif"))
writeRaster(alrM_aoa$AOA, paste0("output/mud_aoa_", Sys.Date(), ".tif"))
```


## Output a log file

```{r log}
sink(file = paste0("output/ModelLog_", Sys.Date(), ".txt"))
print("Selected Predictors")
sel_preds
model
print("Final Model")
paste0("MSE = ", validation[1,1])
paste0("RMSE = ", validation[1,2])
paste0("R2 = ", validation[1,3])
paste0("AOA = ", round(cellStats(alrM_aoa$AOA, stat = "mean"), 2))
sink()
```


# Finishing off


# Save RF model

```{r save_model}
saveRDS(model, "rfmodel.rds")
```


## Save session info

```{r save_session_info}
sessionInfo <- sessionInfo()
save(sessionInfo, file = "sessionInfo.Rdata")
rm("sessionInfo")
```


## Save global environment

```{r save_global_env}
save.image(file = "globEnv.RData")
```
